Changes in gene expression associated with transitions in reproductive meristem identity during early development of rice inflorescence.

### 2015-12-10 ms.Rmd
References

Compared to the switch from apical to axillary meristem, more gradual changes in gene expression were apparent during the sequential transition from PBM to SM. The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` corresponds with their involvement in flower development in rice and other plants. In the eudicot *A. thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity, and members of the *AGL2*-like clade interact with other MADS genes to specify floral organ identity, such as *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), which combine with *APETALA2* (*AP2*) to specify petals and another MADS gene, *AGAMOUS* (*AG*), to specify stamen [recently reviewed by @Prunet:2013ki]. In rice, functions similar to those provided by *A. thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [**ref Nagasawa 2003**; **ref Yao 2008**], whilst stamen and carpels are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [**ref kyozuka 2002**; **ref Yamaguchi 2006**]. Two recent studies in *A. thaliana* and the domesticated tomato, *Solanum lycopersicum*, used LMD and RNA sequencing to study inflorescence development [@Park:2012dg; @Mantegazza:2014di]. Comparison of these datasets with the rice LMD dataset presented here suggests some differences in expression of MIKC^C^ genes between these species (`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`). *A. thaliana* *AGL6* is the only gene from the *AGL6*-like clade that is more highly expressed in indeterminate meristems than determinate meristems. *AGL6* mutants have fewer axillary buds on the inflorescence stem, suggesting that *AGL6* promotes axillary meristem formation [@Huang:2012de]. The rice genes from the *AGL6*-like clade, *MFO* (*MADS6*) and *MADS17*, redundantly promote floral meristem determinancy [@Ohmori:2009kj], and are correspondingly enriched in determinate meristems. Expression of the putatative tomato *AGL6*-like gene *SOLYC01G093960.2* is also enriched in determinate meristems, suggesting that the role of *AGL6* in axillary meristem formation may be specific to the Arabidopsis lineage. In contrast, consistent with their functions in floral organ specification in rice and *A. thaliana*, genes in the *AGL2*-like clade are generally more highly expressed in determinate meristems.

@@

*log1* mutants also appear to have reduced *HOMEOBOX 1* (*OSH1*) expression [@Kurakawa:2007go]. *OSH1* encodes a class I *Knotted1-like homeobox* (*KNOX*) homeodomain transcription factor. The expression of class I *KNOX* genes in young inflorescences has been detected by *in situ* hybridisation [**ref Sentoku 1999**]. *osh1* mutants have defects in SAM maintenance during vegetative growth and lower expression of other *KNOX* genes, and expression of *KNOX* genes can be induced by cytokinin treatment during shoot regeneration [@Tsuda:2011gz]. Another class I *KNOX* gene, *HOMEOBOX 15* (*OSH15*) is expressed in a similar pattern to *OSH1* (`r I(pasteLabel("Figure ", figCount, "f_hb"))`), as previously reported in *O. sativa* cv. Nipponbare [**ref Sato 1998**]. OSH1 binds to the upstream region of the *OSH15* locus, and *osh1 osh15* double-mutants have lower induction of *OSH6* after treatment with cytokinin [@Tsuda:2011gz], supporting a role of *KNOX* genes in the response to phytohormone, and the higher expression of *OSH1* and *OSH15* in indeterminate axillary meristems is compatible with a function in the promotion of BM identity. In *A. thaliana*, cytokinin increases the production of the homeodomain gene *WUSCHEL*, which itself enhances cytokinin signalling [**ref Leibfried 2005**; **ref Lindsay, 2006**; @Gordon:2009fp]. The rice ortholog of *WUSCHEL*, *WUSCHEL-LIKE HOMEOBOX 1* (*WOX1*/*WUS*) [**ref Nardmann Werr 2006**], was not detected in the meristem tissues described here, but another *WUSCHEL homeobox* (*WOX*) gene, *WOX8*, has a peak in expression in RM, and a third, *LOC_Os01g70810*, is predominantly expressed in the RM and PBM (`r I(pasteLabel("Figure ", figCount, "f_hb"))`). Although these results and previous studies support a role of *KNOX* and *WOX* genes in the maintenance of indeterminate meristem identity, possibly coordinated by phytohormone signalling, more work is required to elucidate this mechanism.

Three class IV HD-Zip genes, *RICE OUTMOST CELL-SPECIFIC GENE 1* (*ROC1*), *ROC3* and *LOC_Os09g35760*, were recovered in cluster `r getClusterName('inc', 'small')`, containing genes that had lowest expression in the RM and higher expression in the samples from axillary meristems. A prominent group of homeobox genes follow a similar pattern, including five of the eight detected class IV HD-Zip genes and several genes from other homeobox subclasses, including *OSH1* (`r I(pasteLabel("Figure ", figCount, "f_hb"))`). *Zea mays* class IV HD-Zip genes are expressed in the outer cell layer of the SAM and overexpression of *OUTER CELL LAYER 1* delays flowering in transgenic plants [**ref Javelle 2011, Depege-Fargeix 2011**]. In rice, several class IV HD-Zip genes have been detected in the vegetative SAM by *in situ* hybridisation [**ref Ito 2003**], but their role in axillary meristems in the developing inflorescence has not been explained. There are also smaller groups of homeobox genes that peak in expression in the SM or RM samples, suggesting multiple roles and possible redundancy for homeodomain proteins during inflorescence development.

>>>>

Compared to the switch from apical to axillary meristem, more gradual changes in gene expression were apparent during the sequential transition from PBM to SM. The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` corresponds with their involvement in flower development in rice and other plants. In the eudicot *A. thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity, and members of the *AGL2*-like clade interact with other MADS genes to specify floral organ identity, such as *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), which combine with *APETALA2* (*AP2*) to specify petals and another MADS gene, *AGAMOUS* (*AG*), to specify stamen [recently reviewed by @Prunet:2013ki]. In rice, functions similar to those provided by *A. thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [@Nagasawa:2003gu; @Yao:2008dy], whilst stamen and carpels are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [@Kyozuka:2002vm; @Yamaguchi:2006br]. Two recent studies in *A. thaliana* and the domesticated tomato, *Solanum lycopersicum*, used LMD and RNA sequencing to study inflorescence development [@Park:2012dg; @Mantegazza:2014di]. Comparison of these datasets with the rice LMD dataset presented here suggests some differences in expression of MIKC^C^ genes between these species (`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`). *A. thaliana* *AGL6* is the only gene from the *AGL6*-like clade that is more highly expressed in indeterminate meristems than determinate meristems. *AGL6* mutants have fewer axillary buds on the inflorescence stem, suggesting that *AGL6* promotes axillary meristem formation [@Huang:2012de]. The rice genes from the *AGL6*-like clade, *MFO* (*MADS6*) and *MADS17*, redundantly promote floral meristem determinancy [@Ohmori:2009kj], and are correspondingly enriched in determinate meristems. Expression of the putatative tomato *AGL6*-like gene *SOLYC01G093960.2* is also enriched in determinate meristems, suggesting that the role of *AGL6* in axillary meristem formation may be specific to the Arabidopsis lineage. In contrast, consistent with their functions in floral organ specification in rice and *A. thaliana*, genes in the *AGL2*-like clade are generally more highly expressed in determinate meristems.

@@

*log1* mutants also appear to have reduced *HOMEOBOX 1* (*OSH1*) expression [@Kurakawa:2007go]. *OSH1* encodes a class I *Knotted1-like homeobox* (*KNOX*) homeodomain transcription factor. The expression of class I *KNOX* genes in young inflorescences has been detected by *in situ* hybridisation [@Sentoku:1999ti]. *osh1* mutants have defects in SAM maintenance during vegetative growth and lower expression of other *KNOX* genes, and expression of *KNOX* genes can be induced by cytokinin treatment during shoot regeneration [@Tsuda:2011gz]. Another class I *KNOX* gene, *HOMEOBOX 15* (*OSH15*) is expressed in a similar pattern to *OSH1* (`r I(pasteLabel("Figure ", figCount, "f_hb"))`), as previously reported in *O. sativa* cv. Nipponbare [@Sato:1998tp]. OSH1 binds to the upstream region of the *OSH15* locus, and *osh1 osh15* double-mutants have lower induction of *OSH6* after treatment with cytokinin [@Tsuda:2011gz], supporting a role of *KNOX* genes in the response to phytohormone, and the higher expression of *OSH1* and *OSH15* in indeterminate axillary meristems is compatible with a function in the promotion of BM identity. In *A. thaliana*, cytokinin increases the production of the homeodomain gene *WUSCHEL*, which itself enhances cytokinin signalling [@Leibfried:2005hw; @Lindsay:2006bh; @Gordon:2009fp]. The rice ortholog of *WUSCHEL*, *WUSCHEL-LIKE HOMEOBOX 1* (*WOX1*/*WUS*) [@Nardmann:2006fd], was not detected in the meristem tissues described here, but another *WUSCHEL homeobox* (*WOX*) gene, *WOX8*, has a peak in expression in RM, and a third, *LOC_Os01g70810*, is predominantly expressed in the RM and PBM (`r I(pasteLabel("Figure ", figCount, "f_hb"))`). Although these results and previous studies support a role of *KNOX* and *WOX* genes in the maintenance of indeterminate meristem identity, possibly coordinated by phytohormone signalling, more work is required to elucidate this mechanism.

Three class IV HD-Zip genes, *RICE OUTMOST CELL-SPECIFIC GENE 1* (*ROC1*), *ROC3* and *LOC_Os09g35760*, were recovered in cluster `r getClusterName('inc', 'small')`, containing genes that had lowest expression in the RM and higher expression in the samples from axillary meristems. A prominent group of homeobox genes follow a similar pattern, including five of the eight detected class IV HD-Zip genes and several genes from other homeobox subclasses, including *OSH1* (`r I(pasteLabel("Figure ", figCount, "f_hb"))`). *Zea mays* class IV HD-Zip genes are expressed in the outer cell layer of the SAM and overexpression of *OUTER CELL LAYER 1* delays flowering in transgenic plants [@Javelle:2011dz; @DepegeFargeix:2011ds]. In rice, several class IV HD-Zip genes have been detected in the vegetative SAM by *in situ* hybridisation [@Ito:2003ta], but their role in axillary meristems in the developing inflorescence has not been explained. There are also smaller groups of homeobox genes that peak in expression in the SM or RM samples, suggesting multiple roles and possible redundancy for homeodomain proteins during inflorescence development.
